---
title: "Reproducible Research - NOAA Storm Database"
author: "EJ"
date: "5/26/2020"
output: html_document
---

## Questions to answer
1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Load Data into R
Load the dataset into R and view basic data information

```{r}
StormData <- read.csv("repdata_data_StormData.csv")
dim(data)
colnames(data)
```

## Clean Data

Event titles have multinames for the same event. Clearn data to have unified event names.

```{r}
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData$EVTYPE[grepl("/|&|and", StormData$EVTYPE,ignore.case = TRUE)] <- "Multiple Event"
StormData$EVTYPE[grepl("volc", StormData$EVTYPE,ignore.case = TRUE)] <- "Volcano"
StormData$EVTYPE[grepl("wind|wnd", StormData$EVTYPE,ignore.case = TRUE)] <- "Wind"
StormData$EVTYPE[grepl("funnel|tornado", StormData$EVTYPE,ignore.case = TRUE)] <- "Tornado"
StormData$EVTYPE[grepl("glaze", StormData$EVTYPE,ignore.case = TRUE)] <- "Glaze"
StormData$EVTYPE[grepl("hail", StormData$EVTYPE,ignore.case = TRUE)] <- "Hail"
StormData$EVTYPE[grepl("dust", StormData$EVTYPE,ignore.case = TRUE)]  <- "Dust"
StormData$EVTYPE[grepl("flood", StormData$EVTYPE,ignore.case = TRUE)] <- "Flood"
StormData$EVTYPE[grepl("ic(e|y)", StormData$EVTYPE,ignore.case = TRUE)] <- "Ice"
StormData$EVTYPE[grepl("fire|smoke", StormData$EVTYPE,ignore.case = TRUE)] <- "Fire"
StormData$EVTYPE[grepl("thunder", StormData$EVTYPE,ignore.case = TRUE)] <- "Thunder Storm"
StormData$EVTYPE[grepl("slide|eros", StormData$EVTYPE,ignore.case = TRUE)] <- "Erosion"
StormData$EVTYPE[grepl("rain", StormData$EVTYPE,ignore.case = TRUE)] <- "Rain"
StormData$EVTYPE[grepl("freez|cold|snow|chill|winter", StormData$EVTYPE,ignore.case = TRUE)] <- "Cold Weather"
StormData$EVTYPE[grepl("TROPICAL.STORM", StormData$EVTYPE,ignore.case = TRUE)] <- "Tropical Store"
StormData$EVTYPE[grepl("heat", StormData$EVTYPE,ignore.case = TRUE)] <- "Heat"
StormData$EVTYPE[grepl("(hurri|opal)", StormData$EVTYPE,ignore.case = TRUE)] <- "Hurricane"
```

## Isolate Needed Data

Create two unique sub-datasets for health and economy

```{r}
StormData_Health <- StormData[,(c(8,23:24))]
StormData_Economy <- StormData[,c(8,25:28)]
colnames(StormData_Health)
colnames(StormData_Economy)
```

## Focus on health data

Continue to clean data

```{r}
library(dplyr)
StormData_Health_Aggre <- aggregate(cbind(FATALITIES,INJURIES) ~ EVTYPE, data = StormData_Health, sum, na.rm=TRUE)
StormData_Health_Aggre <- arrange(StormData_Health_Aggre, desc(FATALITIES+INJURIES))
StormData_Health_Aggre_top <- StormData_Health_Aggre[1:10,]
```

Visualize results for health data

```{r}
head(StormData_Health_Aggre_top)
library(ggplot2)
qplot(EVTYPE, FATALITIES+INJURIES, data = StormData_Health_Aggre_top, main = "Impact of Weather Events on Health Damange")
```

## Focus on economy data

Clean economy data for analysis

```{r}
StormData_Economy$PROPDMGCALC [StormData_Economy$PROPDMG==0] <- 0  
StormData_Economy$CROPDMGCALC [StormData_Economy$CROPDMG==0] <- 0 
StormData_Economy$PROPDMGCALC [StormData_Economy$PROPDMGEXP=="H"| StormData_Economy$PROPDMGEXP=="h"]<- StormData_Economy$PROPDMG[StormData_Economy$PROPDMGEXP=="H"|StormData_Economy$PROPDMGEXP=="h"]*100
StormData_Economy$CROPDMGCALC [StormData_Economy$CROPDMGEXP=="H"| StormData_Economy$CROPDMGEXP=="h"]<- StormData_Economy$CROPDMG[StormData_Economy$CROPDMGEXP=="H"|StormData_Economy$CROPDMGEXP=="h"]*100
StormData_Economy$PROPDMGCALC [StormData_Economy$PROPDMGEXP=="K"| StormData_Economy$PROPDMGEXP=="k"]<- StormData_Economy$PROPDMG[StormData_Economy$PROPDMGEXP=="K"|StormData_Economy$PROPDMGEXP=="k"]*1000
StormData_Economy$CROPDMGCALC [StormData_Economy$CROPDMGEXP=="K"| StormData_Economy$CROPDMGEXP=="k"]<- StormData_Economy$CROPDMG[StormData_Economy$CROPDMGEXP=="K"|StormData_Economy$CROPDMGEXP=="k"]*1000
StormData_Economy$PROPDMGCALC [StormData_Economy$PROPDMGEXP=="M"| StormData_Economy$PROPDMGEXP=="m"]<- StormData_Economy$PROPDMG[StormData_Economy$PROPDMGEXP=="M"|StormData_Economy$PROPDMGEXP=="m"]*1000000
StormData_Economy$CROPDMGCALC [StormData_Economy$CROPDMGEXP=="M"| StormData_Economy$CROPDMGEXP=="m"]<- StormData_Economy$CROPDMG[StormData_Economy$CROPDMGEXP=="M"|StormData_Economy$CROPDMGEXP=="m"]*1000000
StormData_Economy$PROPDMGCALC [StormData_Economy$PROPDMGEXP=="B"| StormData_Economy$PROPDMGEXP=="b"]<- StormData_Economy$PROPDMG[StormData_Economy$PROPDMGEXP=="B"|StormData_Economy$PROPDMGEXP=="b"]*1000000000
StormData_Economy$CROPDMGCALC [StormData_Economy$CROPDMGEXP=="B"| StormData_Economy$CROPDMGEXP=="b"]<- StormData_Economy$CROPDMG[StormData_Economy$CROPDMGEXP=="B"|StormData_Economy$CROPDMGEXP=="b"]*1000000000

StormData_Economy_Aggre <- aggregate(cbind(PROPDMGCALC,CROPDMGCALC)~EVTYPE, data = StormData_Economy, sum, na.rm=TRUE)
StormData_Economy_Aggre <- arrange(StormData_Economy_Aggre, desc(PROPDMGCALC+CROPDMGCALC))

StormData_Economy_Aggre_top <- StormData_Economy_Aggre[1:10,]
```

Visualize results for health data

```{r}
head(StormData_Economy_Aggre_top)
qplot(EVTYPE, PROPDMGCALC+CROPDMGCALC, data = StormData_Economy_Aggre_top, main = "Impact of Weather Events on Economic Damange")
```
